---
title: Prepare your computer for R
authors: 
  - 'cb'
date: '2022-06-13'
slug: prepare-your-computer-for-r
categories:
  - Handout
output: 
  html_document: 
    toc: yes
    code_folding: show
tags: []
subtitle: ''
summary: ''
lastmod: '2022-06-13T20:38:45-04:00'
featured: yes
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

## AGENDA

1. Introductions
2. Installing R
3. This Worksheet :)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Adpated from material curated by Eugene Plavskin and Grace Avecilla

## Exploring [RStudio](https://www.rstudio.com/products/rstudio/download/)

- The console: Where you run commands one by one (bottom)
- The environment: Where you can see what is loaded in R (upper right)
- The file explorer: An easier way for you to find files (otherwise you'll need to know the _path_ of a file to use it in R; lower right)

## Installing _packages_

- Packages are like kits in experiments: They _package_ the reagents and instructions for otherwise complicated processes for you.

- We'll be installing 3 packages today
  - [rmarkdown](https://rmarkdown.rstudio.com): A tool that allows you to write your code along with your notes
  - [ggplot2](https://ggplot2.tidyverse.org): A handy and powerful visualization tool
  - [dplyr](https://dplyr.tidyverse.org): A tool that handles and summarizes tables
  
```{r eval = FALSE}
# This is how we ask R to install packages
# R will try to ask several central repositories if they have this package
install.packages("rmarkdown")
```


## Using Rmarkdown

Rmarkdown file allow you to record your thinking process (as text) along with 
your analyses (as _code chunks_).

Having why you are doing the analysis and the specific code for it will help 
you catch up faster if you ever need to go back to the same analysis again.

### Text and code

This is a line of text:

"I know how to use R!"

```{r}
# This is a chunk of code. Do you notice that the background looks different?
# Also there is three buttons at the upper right corner

# Press the _play_ button, and the output will be shown below the chunk
print("I know how to use R")

2+2
```


- Running code:
 - Click on the run button on the upper-right corner of a chunk
 - Ctrl + Enter (Windows) / Cmd + Return (Mac)
 
- Generating output
  - Click the *knit* button on the upper rim of the editor
  - Knitting will _re-run all your code in the file from scratch_, which is a 
  good thing because it makes sure:
    - The code works on its own
    - We haven't accidentally deleted or added some lines of code that will 
    break the code.

## Working directory and Rstudio projects

### What is a working directory?

A working directory is where R will try to find things if you ask it to[^1].

### Checking and setting your working directory

You can check the path of your working directory with `getwd()` 
(short for _get working directory_).

```{r}
# Expect to see a path printed below
# This is where R find things for you
getwd()
```
  
### Loading files from your working directory

Now that you have a directory, files inside that directory can be loaded using 
functions like `read.csv()`. This function will take a comma-separated file 
(which you can make in Excel) and allow you to save it as a variable that R can 
work with.

For the example below, `mytestdata.csv` can be found in the [example data](assets/session1.zip).

```{r}
# Open the csv file with Excel or Google Sheet to take a look
# mytestdata

# And then we load it to see how it looks like in R
read.csv("mytestdata.csv")

# You need to ask R to keep the csv files
# *Check the environment tab on the upper right corner of your window
mytestdata <- read.csv("mytestdata.csv")

# You can ask R to retrieve what it keeps later by the name you gave to the data
mytestdata
```


## Basic usage of R: Using it as a calculator

Enter the following line of command in the console

```{r}
17 + 2.3
```

In programming, division is often written as slashes (`/`), while 
multiplication is written as asterisks (`*`). Now, let's try doing these in the 
console as well:

```{r}
# Try multiplying 8 by 7

8*7
```


```{r}
# Try dividing 48 by 12

48/12
```


## An example of our data

Cassandra's data is allele frequencies as read counts from sequencing. To 
analyze it, she first:

1. loads in the data,
2. runs an R function to analyze it,
3. plots using `ggplot2`.

Here is a quick version of this:


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

#Read in the data
WeightedMeanZStats.100 <- read.csv("CBData.csv")

#Look at the data (this only prints the first 6 rows)
head(WeightedMeanZStats.100)

#Find the dimensions of the datast
dim(WeightedMeanZStats.100)

unique(WeightedMeanZStats.100$Stat)

#Plot the data using ggplot (which we'll go into more later!)
WeightedMeanZStats.100 %>% filter(Stat == "ZPrime") %>% pivot_longer(c("Bulk", "Parent", "Int")) %>% 
  ggplot(., aes(x = POS, y = value, color = name)) + geom_line(size = 1) + #geom_point(alpha = 0.1) + #+ 
  geom_hline(yintercept = 0, color = "gray") +
  geom_hline(yintercept = c(1.96, -1.96), color = "black", linetype = "dashed") +
  theme(axis.text.x = element_blank()) +  ylab("Oak <- Effect -> Wine") +
  scale_color_manual(values = c("violet", "lightblue", "darkorange")) +
  facet_grid(~CHROM, scales = "free") + ggtitle("Comparison of Zprime per Factor | Weighted Means W = 100")

```

[^1]: See [working directory](/2022/06/14/working-directories) if you are interested.
