---
title: 'Session 3: Hands-On Data Wrangling'
authors: ['ycc']
date: '2022-06-21'
slug: session-3-hands-on-data-wrangling
categories:
  - Handout
output:
  html_document:
    toc: yes
    code_folding: show
    code_download: true
tags: []
subtitle: ''
summary: ''
lastmod: '2022-06-21T10:37:17-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
## Loading data

Following what we did last week, we are going to keep working with the penguins 
today. You can find the dataset after you extracted `session3.zip` which 
you can download from [here](https://nyusurp.github.io/assets/session3.zip).

### Figure things out on the internet

If you don't remember how to do something with R and happen to be the most 
R-fluent people in the peer, don't panic. Most of the time, searching for 
_[what you want to do] in R_ work out great.

Online forums like StackOverflow and Kaggle often give great answers with code 
examples that you could play with. If you are a genomic person, Biostar and 
SEQanswers would be your friend.

```{r Loading-data}
library(dplyr)
# How do you load a csv file with R?
penguins <- read.csv("penguins.csv")
```

### Figuring things out within R

Oftentimes, an answer you found online provides a code that almost works, but 
you might not yet know enough to make it work. For times like this, R has 
documentation built in for each function describing the arguments that you can 
tweak and what it meant.

You can trigger the help page with `?[function]` in the console. Let's try it 
out: We used `head()` last week to take a glimpse of the first few rows of a 
table, so how do we print the first _3_ instead of 6 rows?

```{r First-3-rows}
head(penguins, n = 3)
```

## `count()` things

How many penguins of each `species` were observed on each `island`?

```{r count-per-species-per-island}
penguins %>%
  count(island, species)
```

Which species is found on all islands? I know it's obvious, but let's use a 
sledgehammer to crack a nut this time.

```{r most-prevalent-species}
# The output in the previous chunk is also a table that could be count()ed.
penguins %>%
  count(island, species) %>%
  count(species)
```

## `filter()` out your species

Let's say you are interested in comparing how the islands influences the growth 
of penguins. Either you'll need to go to Antarctica to observe more Gentoo or 
Chinstrap penguins on other islands, or only Adelie penguins make sense for 
your purpose.

```{r subset-Adelie-penguins}
# Since none of us is going to leave for Antarctica any time soon (right?)
# Let's keep only the Adelie penguins, and **assign it to a new object**.
adelie <- penguins %>%
  filter(species == "Adelie")
```

### Revisiting the criteria for `filter()`

There are 6 basic types of comparison:

- `==`: Equal to
- `!=`: Not equal to
- `>`: Larger than
- `>=`: Larger or equal to
- `<`: Less than
- `<=`: Less than or equal to

```{r filter-with-equal}
# How many male Adelie penguins were observed?
adelie %>%
  filter(sex == "male") %>%
  count()
```

```{r filter-with-not-equal}
# How many Adelie penguins were found on islands that are not Dream island?
adelie %>%
  filter(island != "Dream") %>%
  count()
```

```{r filter-less-or-equal}
# How many Adelie penguins were observed during or before 2008?
adelie %>%
  filter(year <= 2008) %>%
  count()
```

### Combining criteria

Sometimes, we'll need more than one criteria to get the data we want.
For example, if you are interested in the female Adelie penguins on the Biscoe 
island:

```{r two-filter-criteria}
adelie %>%
  filter(sex == "female", island == "Biscoe")
```

What if you need penguins with extreme body weight? Say either over 4700g or 
below 2900g.

```{r or-gate}
adelie %>%
  filter(body_mass_g > 4700 | body_mass_g < 2900)
```

#### Logical operations: What's happening under the `dplyr` table

Essentially, what `filter()` relies on is to ask a series of yes/no question 
to each row of a column, and the answers to these yes/no questions are called 
Boolean or logical values.

In R, `TRUE` means yes, while `FALSE` means no.

```{r logical-1}
# Is 3 larger than 5?
3 > 5
```

```{r logical-2}
# Is 2022 equal to 2020?
2022 == 2020
```

```{r logical-3}
# Is "apple" not equal to "orange"?
"apple" != "orange"
```

These logical values can be further calculated with `&` (AND), `|` (OR), and 
`xor()`.

|            |  `&`  |  `|`  |`xor()`|
|------------|:-----:|:-----:|:-----:|
|  TRUE/TRUE | TRUE  | TRUE  | FALSE |
| TRUE/FALSE | FALSE | TRUE  | TRUE  |
| FALSE/TRUE | FALSE | TRUE  | TRUE  |
| FALSE/FALSE| FALSE | FALSE | FALSE |

So, how many Adelie penguins either have a bill length shorter than 35mm 
or a bill depth below 15mm?

```{r bill-or-practice}
adelie %>%
  filter(bill_length_mm < 35 | bill_depth_mm < 15) %>%
  count()
```

Now that we know `filter()` is working with logical values, you won't be 
surprised that it can also use other functions that gives a logical value 
as its output as a criterion.

For example, we mentioned that `NA` is how R labels missing data. Since we are 
interested in body weight, we might want throw away rows with missing body 
weight...

```{r NA-is-a-special-case}
# You might want to try filtering for body_mass_g== NA
adelie %>%
  filter(body_mass_g == NA)
```

What is happening? Remember that we said R treats NA very differently. As a 
matter of fact, almost every operation gives you `NA` when `NA` is involved.

```{r NA-is-I-dont-know}
# Try these
5 == NA
3 > NA
"North America" != NA
3.1415926 <= NA
```

Since `NA` is missing data, this actually makes sense:

`5 == NA` is like asking "is 5 equal to something I don't know?", and the 
answer has to be "I don't know".

So, how do we ask R if it doesn't know something or has an `NA` there? We use 
`is.na()`.

```{r try-isna}
is.na(NA)
is.na("National Academy")
```

You can use `is.na()` with `filter()` to find the rows where body weight is 
`NA`.

```{r filter-na}
# Use is.na() to get rows with NA in body_mass_g
adelie %>%
  filter(is.na(body_mass_g))
```

How do we select all other rows then? In R, `!` flips a logical value, and you 
can pronounce it as _not_.

```{r not-logical}
# Not TRUE
!TRUE
# Not FALSE
!FALSE
# Flipping (3 > 5)
!(3 > 5)
```

We can use `!` to find rows that are _not_ `NA`.

```{r drop-nas}
# How many Adelie penguins have their body mass recorded?
adelie %>%
  filter(!is.na(body_mass_g)) %>%
  count()
```

### Extended reading: Dealing with missing values

Depends on what you work with, you might not always want to drop missing values.

If you are not going to throw those data away, you'll need to ask yourself 
questions like "Does missing value appear randomly?", and decide how to best 
deal with them.

If you are interested, Hadley Wickham has 
[a section in his book _R for Data Science_](https://r4ds.hadley.nz/missing-values.html)
discussing this.

## Group your data for further analysis

Let's say we are interested in how body weight differs between islands.
To perform this analysis, we might want to calculate the median and standard 
deviation of body mass per island.

To indicate how `dplyr` should group your data, we use `group_by()` to tell 
it which column contains the group labels.

Let's assign the grouped data to another object named `adelie_per_island`

```{r group-by-island}
adelie_per_island <- adelie %>%
  filter(!is.na(body_mass_g)) %>%
  group_by(island)
```

## Get summary statistics with `summarize()`

Now that we grouped the data, we can calculate the summary statistics.

The syntax for `summarize()` is as follows:

```{r mean-exp}
# You tell summarize() which functions to use on which column
adelie_per_island %>%
  summarize(mean(body_mass_g))
```

By default, the column names of the summary table is the function call we gives
to `summarize()`, which could be a bit ugly. We can rename the columns by:

```{r mean-exp-with-cname}
adelie_per_island %>%
  summarize(mean_bw = mean(body_mass_g))
```

The summary statistics that we use the most often are:

- Arithmetic mean (`mean()`)
- Median (`median()`)
- Standard deviation (`sd()`)

Let's summarize the data with median and standard deviation per island.

```{r summarize-practice}
# Summarize the Adelie subset per island with the median and standard deviation 
# of the body mass
adelie_per_island %>%
  summarize(bw_median = median(body_mass_g), bw_sd = sd(body_mass_g))
```

### What if you don't group?

```{r try-without-grouping}
adelie %>%
  filter(!is.na(body_mass_g)) %>%
  summarize(bw_median = median(body_mass_g), bw_sd = sd(body_mass_g))
```

### What if you want to group with another grouping variable?

```{r alt-group}
# Let's group with sex
adelie %>%
  filter(!is.na(body_mass_g), !is.na(sex)) %>%
  group_by(sex) %>%
  summarize(bw_median = median(body_mass_g), bw_sd = sd(body_mass_g))
```

## A picture is worth a thousand words

While it's great to have summary statistics, oftentimes a quick visualization 
of our data will be very helpful.

`qplot()` provided by `ggplot2` is designed for this purpose and can be easily 
incorporated into your pipeline.

### Example: a quick scatter/strip plot per island

Let's plot something with _points_ on a plane where the x-axis is the islands 
and the y-axis is the body weight.

```{r qplot-strip}
# Load ggplot2
library(ggplot2)

adelie_per_island %>%
  qplot(data = ., x = island, y = body_mass_g, geom = "point")
```

There several required arguments for `qplot()`:

- data: The data you want to plot.
- x: The column name of the variable you want to use as the x-axis
- y: The column name of the variable you want to use as the y-axis
- color: The column name of the variable you want to color your plot by
- geom: Short for "geometry", the type of plots you want to do. Popular ones 
include:
  - point
  - bar
  - boxplot
  - violin
  - histogram
  - density
  
```{r qplot-box}
# Let's do a boxplot instead
adelie_per_island %>%
  qplot(data = ., x = island, y = body_mass_g, geom = "boxplot")
```
### What is the distribution of bill lengths of the whole dataset?

```{r qplot-histogram}
# Plot a histogram with bill lengths on the x-axis
penguins %>%
  qplot(data = ., x = bill_length_mm, geom = "histogram")
```

### Is bill length and depth correlated for the Adelie penguins?

```{r qplot-scatter}
# Plot a point plot with bill length on the x, bill depth on the y, and
# color by sex
adelie %>%
  qplot(
    data = .,
    x = bill_length_mm,
    y = bill_depth_mm,
    color = sex,
    geom = "point"
  )

```

